Comparing the Effectiveness of the Choropleth Map with a Hexagon Tile Map for Communicating Cancer Statistics

### 2023-03-14 paper/paper.Rmd
updated to use thyroid

---
title: "Comparing the Effectiveness of the Choropleth Map with a Hexagon Tile Map for Communicating Cancer Statistics"
affiliation:
  author-columnar: true         ## one column per author
  #institution-columnar: true  ## one column per institution (multiple autors eventually)
  # wide: true                  ## one column wide author/affiliation fields
  institution:
    - name: Queensland University of Technology
      department: Science and Engineering Faculty
      location: Brisbane, Australia
      email: stephanie.kobakian@qut.edu.au
      mark: 1
      author:
        - name: Stephanie Kobakian
    - name: Monash University
      department: Econometrics and Business Statistics Faculty
      location: Melbourne, Australia
      email: dicook@monash.edu
      mark: 1
      author:
        - name: Dianne Cook
keywords: ["statistics", "visual inference", "geospatial", "population"]
abstract: |
  The choropleth map display is commonly used for communicating spatial distributions across geographic areas. However, when choropleths are used the size of the geographic units will influence the understanding of the distribution derived by map users. The hexagon tile map is presented as an alternative display for visualizing population related distributions effectively. Visual inference is used to measure the power of the hexagon tile map design, and the choropleth is used as a comparison. The hexagon tile map display is tested using a distribution that is directly related to the geography, with values monotonically increasing from the North-West to South-East areas of Australia. This study finds in a hexagon tile map lineup the single map that contains a population related distribution is detected with greater probability than the same data displayed in a choropleth map. These findings should encourage map creators to implement alternative displays and consider a hexagon tile map when presenting spatial distributions of heterogeneous areas.
dev: pdf
bibliography: paper.bib
output:
  bookdown::pdf_book: 
      base_format: rticles::ieee_article
      number_sections: true
editor_options: 
  chunk_output_type: inline
link-citations: true
citation_package: biblatex
csl: springer-vancouver.csl
header-includes: \usepackage{graphicx}
---

@@

@@

```{r libraries}
# Load Libraries
library(tidyverse)
library(readxl)
library(broom)
library(cowplot)
library(png)
library(grid)
library(lme4)
library(ggthemes)
library(RColorBrewer)
library(knitr)
library(kableExtra)

@@

@@

```{r liver, fig.height = 4, fig.width = 4, fig.cap = "A choropleth map of the smoothed average of liver cancer diagnoses for Australian males. The diverging colour scheme uses dark blue areas for much lower than average diagnoses, yellow areas with diagnoses around the Australian average, red shows diagnoses much higher than average. The hexagon tile map shows concentrations of higher than expected liver cancer rates in the cities of Melbourne and Sydney, which is not visible from the choropleth.", results = "asis"}
ggdraw() +
  draw_plot(rasterGrob(readPNG("figures/aus_liver_m.png")))
```

```{r liver-hex, fig.height = 4, fig.width = 4, fig.cap = "A hexagon tile map of the smoothed average of liver cancer diagnoses for Australian males. The diverging colour scheme uses dark blue areas for much lower than average diagnoses, yellow areas with diagnoses around the Australian average, red shows diagnoses much higher than average. The hexagon tile map shows concentrations of higher than expected liver cancer rates in the cities of Melbourne and Sydney, which is not visible from the choropleth.", results = "asis"}

@@

The choropleth map is an effective spatial display if the size of the geographic units is relatively uniform. This is not the case for most countries. Size heterogeneity in administrative units is particularly extreme in Australia: most of the landscape of Australia is sparsely settled, with the population densely clustered into the narrow coastal strips. A choropleth map focuses attention on the geography, and for heterogeneously sized areas it presents a biased view of the population related distribution of the statistic [@CBATCC]. *Land doesn't get cancer, people do* -- a more effective way to communicate the spatial distributions of cancer statistics is needed.

@@

<!-- Hexagon tile map -->
The hexagon tile map algorithm, automatically matches spatial regions to their nearest hexagon tile, from a grid of tiles. It has the effect of spreading out the inner city areas while maintaining the spatial locations or regions in remote areas. The algorithm is available in the R package, sugarbag [@sugarbag]. Fig. \ref{fig:liver} shows the hexagon tile map, along with the choropleth map of liver cancer rates in Australia. Colour maps from substantially below average (blue) to substantially above average (red) rates. The inner city areas have expanded, making it possible to see the cancer incidence in the small, densely populated areas. Remote regions are represented by isolated hexagons, which is not ideal, but maintains the spatial location of these data values. It is of interest to know how well the spatial distribution is perceived for this display, in comparison to the choropleth.

@@

The primary factor in the experiment is the plot type. The secondary factor is a trend model. Three trend models were developed, one mirroring a large spatial trend for which the choropleth would be expected to do well, and two with differing level of inner city hot spots. These latter two reflect the structure seen in the liver cancer data (Fig. \ref{fig:liver}). This produces six treatment levels:

@@

This smoothing allowed neighbors to be related to each other, but also allowed outliers, and showed distributions similar to the Liver cancer distribution (Fig. \ref{fig:liver}).

@@

# Plot summary
ggplot(d_smry, aes(x = type, y = pdetect, color = trend)) +
  geom_point(size = 2) +  
  geom_line(size = 1, aes(group = replicate)) +  
  facet_wrap(~trend) +
  scale_color_manual(values = trend_colors) +
  xlab("Type of areas visualized") +
  ylab("Detection rate") + 
  ylim(0,1) +
  guides(color = FALSE)
```

@@

The source code to produce this document can be found on [GitHub](https://github.com/srkobakian/experiment/tree/master/paper).
Supplementary materials have been included to discuss the survey procedures and the lineups that were used.
The full set of images can also be found at: https://github.com/srkobakian/experiment/tree/master/figures/final.

@@

\newpage
References {.numbered}
==========

>>>>

---
title: "Comparing the Effectiveness of the Choropleth Map with a Hexagon Tile Map for Communicating Cancer Statistics"
affiliation:
  author-columnar: true         ## one column per author
  #institution-columnar: true  ## one column per institution (multiple autors eventually)
  # wide: true                  ## one column wide author/affiliation fields
  institution:
    - name: Queensland University of Technology
      department: Science and Engineering Faculty
      location: Brisbane, Australia
      email: stephanie.kobakian@qut.edu.au
      mark: 1
      author:
        - name: Stephanie Kobakian
    - name: Monash University
      department: Econometrics and Business Statistics Faculty
      location: Melbourne, Australia
      email: dicook@monash.edu
      mark: 1
      author:
        - name: Dianne Cook
keywords: ["statistics", "visual inference", "geospatial", "population"]
abstract: |
  The choropleth map display is commonly used for communicating spatial distributions across geographic areas. However, when choropleths are used the size of the geographic units will influence the understanding of the distribution derived by map users. The hexagon tile map is presented as an alternative display for visualizing population related distributions effectively. Visual inference is used to measure the power of the hexagon tile map design, and the choropleth is used as a comparison. The hexagon tile map display is tested using a distribution that is directly related to the geography, with values monotonically increasing from the North-West to South-East areas of Australia. This study finds in a hexagon tile map lineup the single map that contains a population related distribution is detected with greater probability than the same data displayed in a choropleth map. These findings should encourage map creators to implement alternative displays and consider a hexagon tile map when presenting spatial distributions of heterogeneous areas.
dev: pdf
bibliography: paper.bib
output:
  bookdown::pdf_book: 
      base_format: rticles::ieee_article
      number_sections: true
editor_options: 
  chunk_output_type: console
link-citations: true
citation_package: biblatex
csl: springer-vancouver.csl
header-includes: \usepackage{graphicx}
---

@@

<!-- Changes: thyroid instead of liver - check; lineup need to be hexmap; check language with chatgpt/grammarly; get refs formatted and paper into 10 pages -->

@@

```{r libraries}
# Load Libraries
library(tidyverse)
library(readxl)
library(broom)
library(broom.mixed)
library(cowplot)
library(png)
library(grid)
library(lme4)
library(ggthemes)
library(RColorBrewer)
library(knitr)
library(kableExtra)

@@

```{r makethyroidplots}
library(cartogram)
library(sugarbag)
library(sf)

# Spatial polygons files sourced from: https://github.com/wfmackey/absmapsdata
load("data/sa22011.rda")
load("data/state2011.rda")

invthm <- theme_map() + 
  theme(
    panel.background = element_rect(fill = "transparent", colour = NA), 
    plot.background = element_rect(fill = "transparent", colour = NA),
    legend.background = element_rect(fill = "transparent", colour = NA),
    legend.key = element_rect(fill = "transparent", colour = NA),
    text = element_text(colour = "white"),
    axis.text = element_text(colour = "white")
  )

# function to allocate colours to regions
aus_colours <- function(sir_p50){
  value <- case_when(
    sir_p50 <  0.74 ~ "#33809d",
    sir_p50 >= 0.74 & sir_p50 < 0.98 ~ "#aec6c7",
    sir_p50 >= 0.98 & sir_p50 < 1.05 ~ "#fff4bc",
    sir_p50 >= 1.05 & sir_p50 < 1.45 ~ "#ff9a64",
    sir_p50 >= 1.45 ~ "#ff3500",
    TRUE ~ "#FFFFFF")
  return(value)
}

sa2 <- sa22011 %>% 
  filter(!st_is_empty(geometry)) %>% 
  filter(!state_name_2011 == "Other Territories") %>% 
  filter(!sa2_name_2011 == "Lord Howe Island")

SIR <- read_csv("data/SIR Downloadable Data.csv") %>% 
  filter(SA2_name %in% sa2$sa2_name_2011) %>% 
  dplyr::select(Cancer_name, SA2_name, Sex_name, p50) %>% 
  filter(Cancer_name == "Thyroid", Sex_name == "Females")
ERP <- read_csv("data/ERP.csv") %>%
  filter(REGIONTYPE == "SA2", Time == 2011, Region %in% SIR$SA2_name) %>% 
  dplyr::select(Region, Value)
# Alternative maps
# Join with sa2 sf object
sa2thyroid_ERP <- SIR %>% 
  left_join(sa2, ., by = c("sa2_name_2011" = "SA2_name")) %>%
  left_join(., ERP %>% 
              dplyr::select(Region, 
              Population = Value), by = c("sa2_name_2011"= "Region")) %>% 
  filter(!st_is_empty(geometry))
sa2thyroid_ERP <- sa2thyroid_ERP %>% 
  #filter(!is.na(Population)) %>% 
  filter(!sa2_name_2011 == "Lord Howe Island") %>% 
  mutate(SIR = map_chr(p50, aus_colours)) %>% 
  st_as_sf()

# Make choropleth
aus_ggchoro <- ggplot(sa2thyroid_ERP) + 
  geom_sf(aes(fill = SIR), size = 0.1) + 
  scale_fill_identity() + invthm

# Make hexmap
if (!file.exists("data/aus_hexmap.rda")) {

## Check the projection uses long and lat
## Create centroids set
sa2centroids <- sa2 %>% 
  create_centroids(., "sa2_name_2011")
## Create hexagon grid
# For speed, consider if the buffer distance beyond the centroids is necessary
grid <- create_grid(centroids = sa2centroids,
                    hex_size = 0.15,
                    buffer_dist = 4, verbose = TRUE)
## Allocate polygon centroids to hexagon grid points
# This is a time consuming process for a large data set
aus_hexmap <- allocate(
  centroids = sa2centroids,
  hex_grid = grid,
  sf_id = "sa2_name_2011",
  ## same column used in create_centroids
  hex_size = 0.15,
  hex_filter = 8,
  focal_points = capital_cities,
  width = 25,
  verbose = TRUE)
save(aus_hexmap, file = "data/aus_hexmap.rda") 
}

load("data/aus_hexmap.rda")

## Prepare to plot
fort_hex <- fortify_hexagon(data = aus_hexmap,
                            sf_id = "sa2_name_2011",
                            hex_size = 0.3) %>% 
            left_join(sa2thyroid_ERP %>%
            select(sa2_name_2011, SIR, p50))
fort_aus <- sa2thyroid_ERP %>%
  fortify_sfc()
## Make a plot
aus_hexmap_plot <- ggplot() +
  geom_polygon(aes(x = long,  y = lat,  
        group = interaction(sa2_name_2011, polygon)),
               fill = "black",  colour = "darkgrey", 
               linewidth = 0.1, data = fort_aus) +
  geom_polygon(aes(x = long, y = lat, group = hex_id, 
                   fill = SIR), data = fort_hex) +
  scale_fill_identity() +
  invthm + coord_equal()
```

@@

```{r liver, eval=FALSE, fig.height = 4, fig.width = 4, fig.cap = "A choropleth map of the smoothed average of liver cancer diagnoses for Australian males. The diverging colour scheme uses dark blue areas for much lower than average diagnoses, yellow areas with diagnoses around the Australian average, red shows diagnoses much higher than average. The hexagon tile map shows concentrations of higher than expected liver cancer rates in the cities of Melbourne and Sydney, which is not visible from the choropleth.", results = "asis"}
ggdraw() +
  draw_plot(rasterGrob(readPNG("figures/aus_liver_m.png")))
```

```{r liver-hex, eval=FALSE, fig.height = 4, fig.width = 4, fig.cap = "A hexagon tile map of the smoothed average of liver cancer diagnoses for Australian males. The diverging colour scheme uses dark blue areas for much lower than average diagnoses, yellow areas with diagnoses around the Australian average, red shows diagnoses much higher than average. The hexagon tile map shows concentrations of higher than expected liver cancer rates in the cities of Melbourne and Sydney, which is not visible from the choropleth.", results = "asis"}

@@

```{r thyroid-choro, out.height = "30%", fig.height = 3, fig.width = 4, fig.align='center', fig.cap = "A choropleth map of thyroid incidence among females across the Statistical Areas of Australia at Level 2. Blue indicates lower than average and red indicates higher than average incidence. A cluster of high incidence is visible on the east coast."}
aus_ggchoro
```

The choropleth map is an effective spatial display if the size of the geographic units is relatively uniform. This is not the case for most countries. Size heterogeneity in administrative units is particularly extreme in Australia: most of the landscape of Australia is sparsely settled, with the population densely clustered into the narrow coastal strips. Fig. \ref{fig:thyroid-choro} shows the choropleth map of thyroid cancer rates in Australia. The choropleth map focuses attention on the geography, and for heterogeneously sized areas it presents a biased view of the population related distribution of the statistic [@CBATCC]. *Land doesn't get cancer, people do* -- a more effective way to communicate the spatial distributions of cancer statistics is needed.

@@

```{r thyroid-hex, out.height = "30%", fig.height = 3, fig.width = 4, fig.align='center', fig.cap = "A hexagon tile map of female thyroid cancer incidence in Australia, the same data as shown in the choropleth map in Figure 1. The high incidence in several of the metropolitan regions (Brisbane, Sydney and Perth) can now be seen, along with numerous isolated spots."}
aus_hexmap_plot
```

<!-- Hexagon tile map -->
The hexagon tile map algorithm, automatically matches spatial regions to their nearest hexagon tile, from a grid of tiles. It has the effect of spreading out the inner city areas while maintaining the spatial locations or regions in remote areas. The algorithm is available in the R package, sugarbag [@sugarbag]. Fig. \ref{fig:thyroid-hex} shows the hexagon tile map. Colour maps from substantially below average (blue) to substantially above average (red) rates. The inner city areas have expanded, making it possible to see the cancer incidence in the small, densely populated areas. Remote regions are represented by isolated hexagons, which is not ideal, but maintains the spatial location of these data values. It is of interest to know how well the spatial distribution is perceived for this display, in comparison to the choropleth.

@@

The primary factor in the experiment is the plot type. The secondary factor is a trend model. Three trend models were developed, one mirroring a large spatial trend for which the choropleth would be expected to do well, and two with differing level of inner city hot spots. These latter two reflect the structure seen in the thyroid cancer data (Fig. \ref{fig:thyroid}). This produces six treatment levels:

@@

This smoothing allowed neighbors to be related to each other, but also allowed outliers, and showed distributions similar to the thyroid cancer distribution (Fig. \ref{fig:thyroid}).

@@

# Plot summary
ggplot(d_smry, aes(x = type, y = pdetect, color = trend)) +
  geom_point(size = 2) +  
  geom_line(size = 1, aes(group = replicate)) +  
  facet_wrap(~trend) +
  scale_color_manual(values = trend_colors) +
  xlab("Type of areas visualized") +
  ylab("Detection rate") + 
  ylim(0,1) +
  guides(scale = "none")
```

@@

The source code to produce this document can be found on [GitHub](https://github.com/srkobakian/experiment/paper).
Supplementary materials have been included to discuss the survey procedures and the lineups that were used.
The full set of images can be found here, too.

@@

References {.numbered}
==========
